---
title: "2026.1.13 MLH Data Visualization"
output: html_notebook
---

Goal summary data to have to present:
1. 13 yo initiation?
2. 13 yo completion
    - to match initial data, would pull 13yo from each survey year
    - alt, could use 2023 data and use utd by 13 flag
3. 16 yo completion
    - no flag, so have to pull 16yo from each survey year and then use utd flag
Use provider weight for all 

1. 13yo HPV initiation from 2023 data
```{r}
#DATASET: NISTEENPUF23_PDAT
#vARIABLE OF INTEREST: P_U13HPV = INITATION BEFORE 13 BDAY


#1 filter 2023 data
NISTEENPUF23_PDAT <- NISTEENPUF23[ NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 

#2 create inferred birth year
NISTEENPUF23_PDAT$INFBIRTHYR <- NISTEENPUF23_PDAT$YEAR - NISTEENPUF23_PDAT$AGE

#3 IF USING MULTIPLE YEARS, NEED TO DIVIDE PROVIDER WEIGHT BY NUMBER OF YEARS HERE

# We multiply the 0/1 status by the weight to get the weighted success count
P_U13HPV_weighted_sum <- tapply(NISTEENPUF23_PDAT$P_U13HPV * NISTEENPUF23_PDAT$PROVWT_C, 
                           NISTEENPUF23_PDAT$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEENPUF23_PDAT_total_weighted_pop <- tapply(NISTEENPUF23_PDAT$PROVWT_C, 
                             NISTEENPUF23_PDAT$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_U13HPV_START_by_age_23 <- (P_U13HPV_weighted_sum / NISTEENPUF23_PDAT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_U13HPV_START_by_age_23, 
        main = "2023: Percentage of Teens Initiated by 13yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Initiated (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
This reflects previous studies suggesting decreased initiation in 2008 cohort.


2. 13yo HPV completion from 2023 data
```{r}
#DATASET: NISTEENPUF23_PDAT
#vARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY


#1 filter 2023 data
NISTEENPUF23_PDAT <- NISTEENPUF23[ NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 

#2 create inferred birth year
NISTEENPUF23_PDAT$INFBIRTHYR <- NISTEENPUF23_PDAT$YEAR - NISTEENPUF23_PDAT$AGE

#3 IF USING MULTIPLE YEARS, NEED TO DIVIDE PROVIDER WEIGHT BY NUMBER OF YEARS HERE

# We multiply the 0/1 status by the weight to get the weighted success count
P_U13HPV_15INT_weighted_sum <- tapply(NISTEENPUF23_PDAT$P_U13HPV_15INT * NISTEENPUF23_PDAT$PROVWT_C, 
                           NISTEENPUF23_PDAT$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEENPUF23_PDAT_total_weighted_pop <- tapply(NISTEENPUF23_PDAT$PROVWT_C, 
                             NISTEENPUF23_PDAT$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_U13HPV_UTD_by_age_23 <- (P_U13HPV_15INT_weighted_sum / NISTEENPUF23_PDAT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_U13HPV_UTD_by_age_23, 
        main = "2023: Percentage of Teens UTD by 13yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
This reflects previous studies suggesting decreased completion in 2008 cohort.

3. 16 yo completion:have to pull from each survey year

```{r}
#DATASET: NISTEEN_16_PDAT_MLHVARS_ALL
#vARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).


MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3C_15INT")

NISTEENPUF23_16_PDAT <- NISTEENPUF23[NISTEENPUF23$AGE == 16 & NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF22_16_PDAT <- NISTEENPUF22[NISTEENPUF22$AGE == 16 & NISTEENPUF22$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF21_16_PDAT <- NISTEENPUF21[NISTEENPUF21$AGE == 16 & NISTEENPUF21$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF20_16_PDAT <- NISTEENPUF20[NISTEENPUF20$AGE == 16 & NISTEENPUF20$PDAT2 == 1, MLHvars1826 ]
NISTEENPUF19_16_PDAT <- NISTEENPUF19[NISTEENPUF19$AGE == 16 & NISTEENPUF19$PDAT2 == 1, MLHvars1826 ] 

#combine into one dataset
NISTEEN_16_PDAT_MLHVARS_ALL <- rbind(NISTEENPUF23_16_PDAT,NISTEENPUF22_16_PDAT, NISTEENPUF21_16_PDAT, NISTEENPUF20_16_PDAT, NISTEENPUF19_16_PDAT)

# Divide weights by 5 to obtain the revised weights for the 5-year average [6]
NISTEEN_16_PDAT_MLHVARS_ALL$NEWWT <- NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C / 5

# Create inferred birthyear variable
NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR <- NISTEEN_16_PDAT_MLHVARS_ALL$YEAR-16

# We multiply the 0/1 status by the weight to get the weighted success count
P_UTDHPV_15INT_weighted_sum <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$P_UTDHPV_15INT * NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C, 
                           NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C, 
                             NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_UTDHPV_15INT_by_16 <- (P_UTDHPV_15INT_weighted_sum / NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_UTDHPV_15INT_by_16, 
        main = "2019-2023: Percentage of Teens UTD by 16yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
This is where we would want to add in our 2008 16yo data from the 2024 dataset.



