---
title: "2026.01.22 MLH Stats"
output: html_notebook
---

Goal: Add statistics to 16 yo completion rate for presentation. 

16 yo completion
    - no flag, so have to pull 16yo from each survey year and then use utd flag
Use provider weight for all 

Copied below from 2026.1.13 MLH Data Visualization
3. 16 yo completion:have to pull from each survey year

```{r}
#DATASET: NISTEEN_16_PDAT_MLHVARS_ALL
#VARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).


MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3C_15INT", "STRATUM", "P_U13HPV_15INT")

NISTEENPUF23_16_PDAT <- NISTEENPUF23[NISTEENPUF23$AGE == 16 & NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF22_16_PDAT <- NISTEENPUF22[NISTEENPUF22$AGE == 16 & NISTEENPUF22$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF21_16_PDAT <- NISTEENPUF21[NISTEENPUF21$AGE == 16 & NISTEENPUF21$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF20_16_PDAT <- NISTEENPUF20[NISTEENPUF20$AGE == 16 & NISTEENPUF20$PDAT2 == 1, MLHvars1826 ]
NISTEENPUF19_16_PDAT <- NISTEENPUF19[NISTEENPUF19$AGE == 16 & NISTEENPUF19$PDAT2 == 1, MLHvars1826 ] 

#combine into one dataset
NISTEEN_16_PDAT_MLHVARS_ALL <- rbind(NISTEENPUF23_16_PDAT,NISTEENPUF22_16_PDAT, NISTEENPUF21_16_PDAT, NISTEENPUF20_16_PDAT, NISTEENPUF19_16_PDAT)

# Divide weights by 5 to obtain the revised weights for the 5-year average [6]
NISTEEN_16_PDAT_MLHVARS_ALL$NEWWT <- NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C / 5

# Create inferred birthyear variable
NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR <- NISTEEN_16_PDAT_MLHVARS_ALL$YEAR-16

# We multiply the 0/1 status by the weight to get the weighted success count
P_UTDHPV_15INT_weighted_sum <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$P_UTDHPV_15INT * NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C, 
                           NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C, 
                             NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_UTDHPV_15INT_by_16 <- (P_UTDHPV_15INT_weighted_sum / NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_UTDHPV_15INT_by_16, 
        main = "2019-2023: Percentage of Teens UTD by 16yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
This is where we would want to add in our 2008 16yo data from the 2024 dataset.

Now, attempting to perform Logistic Regression for Binary Outcomes. Treat INFBIRTHYR as a categorical factor to determine if specific birth years differ significantly.

Need to incorporate PROVWT_C, STRATUM, AND SEQNUMT.

3.1: Create survey design object.

```{r}
#DATASET: NISTEEN_16_PDAT_MLHVARS_WT
#vARIABLE OF INTEREST: P_UTDHPV_15INT

# 1. Load the survey package (referenced as the standard tool in the sources [11])
library(survey) #i had to manually load this into R

# 2. Define the complex survey design
# Add STRATUM to original dataset....

# Code below was giving me an error bc I had a few NA weights. New dataset filtered for weight not NA:
NISTEEN_16_PDAT_MLHVARS_WT <- NISTEEN_16_PDAT_MLHVARS_ALL[!is.na(NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C), ]

# Use STRATUM for strata and PROVWT_C for weights [14, 15]
PDAT_16_survey <- svydesign(id = ~SEQNUMT, 
                            strata = ~STRATUM, 
                            weights = ~NEWWT, 
                            data = NISTEEN_16_PDAT_MLHVARS_WT, 
                            nest = TRUE)

# 3. Perform Logistic Regression to test for significant differences [8, 11]
# Use 'family = quasibinomial' for survey-weighted binary data
model <- svyglm(P_UTDHPV_15INT ~ as.factor(INFBIRTHYR), 
                design = PDAT_16_survey, 
                family = quasibinomial())

# 4. Review the results for statistical significance (P < 0.05) [6]
summary(model)
```
Ok look at me I got some p-values! Small ones too! This code automatically used 2003 as the reference, 
because it was first in the set. Need to think about what I'm actually looking for for this group...
Evaluating for a trend vs using 2008 as the reference group?

Let's put a pin in that and look at the 13yo HPV completion data for 2023, and try stats on that.

Copied from 2026.1.13 MLH Data Visualization:
2. 13yo HPV completion from 2023 data
```{r}
#DATASET: NISTEENPUF23_PDAT
#vARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY
MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3C_15INT", "STRATUM", "P_U13HPV_15INT")


#1 filter 2023 data
NISTEENPUF23_PDAT <- NISTEENPUF23[ NISTEENPUF23$PDAT2 == 1 & !is.na(NISTEENPUF23$PROVWT_C), MLHvars1826 ] 

#2 create inferred birth year
NISTEENPUF23_PDAT$INFBIRTHYR <- NISTEENPUF23_PDAT$YEAR - NISTEENPUF23_PDAT$AGE

#3 IF USING MULTIPLE YEARS, NEED TO DIVIDE PROVIDER WEIGHT BY NUMBER OF YEARS HERE

# We multiply the 0/1 status by the weight to get the weighted success count
P_U13HPV_15INT_weighted_sum <- tapply(NISTEENPUF23_PDAT$P_U13HPV_15INT * NISTEENPUF23_PDAT$PROVWT_C, 
                           NISTEENPUF23_PDAT$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEENPUF23_PDAT_total_weighted_pop <- tapply(NISTEENPUF23_PDAT$PROVWT_C, 
                             NISTEENPUF23_PDAT$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_U13HPV_UTD_by_age_23 <- (P_U13HPV_15INT_weighted_sum / NISTEENPUF23_PDAT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_U13HPV_UTD_by_age_23, 
        main = "2023: Percentage of Teens UTD by 13yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
This reflects previous studies suggesting decreased completion in 2008 cohort.

Ok...My point is that 2008 is significantly decreased to what it should have been...so significantly less than 2007 and 2009. I mean a lot of them are significantly less than the year after them dt increasing vaccination rates. So might want to think about what we really mean by that.

For now, let's just try to run the stats with 2008 as the reference group.

```{r}
#DATASET: NISTEENPUF23_PDAT
#vARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY

# Use STRATUM for strata and PROVWT_C for weights [14, 15]
PUF23_U13HPV <- svydesign(id = ~SEQNUMT, 
                            strata = ~STRATUM, 
                            weights = ~PROVWT_C, 
                            data = NISTEENPUF23_PDAT, 
                            nest = TRUE)

# 3. Perform Logistic Regression to test for significant differences [8, 11]
# Use 'family = quasibinomial' for survey-weighted binary data
model <- svyglm(P_U13HPV_15INT ~ as.factor(INFBIRTHYR), 
                design = PUF23_U13HPV, 
                family = quasibinomial())

# 4. Review the results for statistical significance (P < 0.05) [6]
summary(model)
```
Okay that had the reference year as 2006, because it was first. Rerun stats but with ref year 2008:

```{r}
#DATASET: NISTEENPUF23_PDAT
#vARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY
library(survey)

# 1. Convert the birth year to a factor
NISTEENPUF23_PDAT$BIRTH_FACTOR <- as.factor(NISTEENPUF23_PDAT$INFBIRTHYR)

# 2. Change the reference level to 2007
# This makes 2007 the new baseline (Intercept)
NISTEENPUF23_PDAT$BIRTH_FACTOR <- relevel(NISTEENPUF23_PDAT$BIRTH_FACTOR, ref = "2008")

# 3. Update your survey design object to include the new variable
PUF23_U13HPV_2008 <- svydesign(id = ~SEQNUMT, 
                            strata = ~STRATUM, 
                            weights = ~PROVWT_C, 
                            data = NISTEENPUF23_PDAT, 
                            nest = TRUE)

# 4. Run the model using the new factor variable
PUF23_U13HPV_2008_model <- svyglm(P_U13HPV_15INT ~ BIRTH_FACTOR, 
                   design = PUF23_U13HPV_2008, 
                   family = quasibinomial())

summary(PUF23_U13HPV_2008_model)
```
This found that 2009 had a statistically significant difference from 2008. Original study shows that 
2007 should also be statistically significant. Likely they did a chi-squared test instead of Fisher.

Ooh I want a graph tracing the rate of vaccination of the 2008s throughout each survey year!

To do that, need to pull 2008 inferred birth year from all datasets. 
2023 = 15yo
2022 = 14yo
2021 = 13yo


```{r}
#DATASET: NISTEEN_08_PDAT_MLHVARS_ALL
#VARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).


MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "STRATUM")

NISTEENPUF23_08_PDAT <- NISTEENPUF23[NISTEENPUF23$AGE == 15 & NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF22_08_PDAT <- NISTEENPUF22[NISTEENPUF22$AGE == 14 & NISTEENPUF22$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF21_08_PDAT <- NISTEENPUF21[NISTEENPUF21$AGE == 13 & NISTEENPUF21$PDAT2 == 1, MLHvars1826 ] 

#combine into one dataset
NISTEEN_08_PDAT_MLHVARS_ALL <- rbind(NISTEENPUF23_08_PDAT,NISTEENPUF22_08_PDAT, NISTEENPUF21_08_PDAT)

# Divide weights by 3 to obtain the revised weights for the 3-year average [6]
NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT <- NISTEEN_08_PDAT_MLHVARS_ALL$PROVWT_C / 3

```

Dataset created, now need to make graph.
```{r}
#DATASET: NISTEEN_08_PDAT_MLHVARS_ALL
#VARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).

# We multiply the 0/1 status by the weight to get the weighted success count
P_UTDHPV_15INT_2008_weighted_sum <- tapply(NISTEEN_08_PDAT_MLHVARS_ALL$P_UTDHPV_15INT * NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT, 
                           NISTEEN_08_PDAT_MLHVARS_ALL$YEAR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEEN_08_PDAT_MLHVARS_ALL_total_weighted_pop <- tapply(NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT, 
                             NISTEEN_08_PDAT_MLHVARS_ALL$YEAR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_UTDHPV_15INT_08 <- (P_UTDHPV_15INT_2008_weighted_sum / NISTEEN_08_PDAT_MLHVARS_ALL_total_weighted_pop) * 100


#4 create bar graph
barplot(P_UTDHPV_15INT_08, 
        main = "2021-2023: % OF 2008 (Inferred) UTD", 
        xlab = "Survey Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 50)) # Sets the axis from 0% to 50%
```

