---
title: "MLH Data for 1.29.26 Meeting"
output: html_notebook
---

1. 2023 Data: Percent of Teens UTD by 13 years old, by birth year. 
Goal: Confirm data from previous studies that suggests 2008 had significantly lower vax rates.

```{r}
#DATASET: NISTEENPUF23_PDAT
#VARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY
MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3C_15INT", "STRATUM", "P_U13HPV_15INT")
#This filters the data to only variables I am interested in. Just allows for a smaller databook w easier processing.
#Note that PU13HPV_15INT is only in the 2023 dataset.

#1 filter 2023 data
NISTEENPUF23_PDAT <- NISTEENPUF23[ NISTEENPUF23$PDAT2 == 1 & !is.na(NISTEENPUF23$PROVWT_C), MLHvars1826 ] 
# filters to teens w adequate provider data, with a PROVWT_C value, and keeps the variables I wanted.

#2 create inferred birth year
NISTEENPUF23_PDAT$INFBIRTHYR <- NISTEENPUF23_PDAT$YEAR - NISTEENPUF23_PDAT$AGE

#3 IF USING MULTIPLE YEARS, NEED TO DIVIDE PROVIDER WEIGHT BY NUMBER OF YEARS HERE

# We multiply the 0/1 status by the weight to get the weighted success count
P_U13HPV_15INT_weighted_sum <- tapply(NISTEENPUF23_PDAT$P_U13HPV_15INT * NISTEENPUF23_PDAT$PROVWT_C, 
                           NISTEENPUF23_PDAT$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEENPUF23_PDAT_total_weighted_pop <- tapply(NISTEENPUF23_PDAT$PROVWT_C, 
                             NISTEENPUF23_PDAT$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_U13HPV_UTD_by_age_23 <- (P_U13HPV_15INT_weighted_sum / NISTEENPUF23_PDAT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_U13HPV_UTD_by_age_23, 
        main = "2023: Percentage of Teens UTD by 13yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```
While these numbers are low, they reflect what is found in the codebook.

1b. Statistics for 2023, UTD by 13 yo, by birth year:

```{r}
#DATASET: NISTEENPUF23_PDAT
#VARIABLE OF INTEREST: P_U13HPV_15INT = UTD BEFORE 13 BDAY
library(survey)
# you may need to download this via Tools -> install packages

# 1. Convert the birth year to a factor (idk why?)
NISTEENPUF23_PDAT$BIRTH_FACTOR <- as.factor(NISTEENPUF23_PDAT$INFBIRTHYR)

# 2. Change the reference level to 2008
# This makes 2008 the new baseline (Intercept)
NISTEENPUF23_PDAT$BIRTH_FACTOR <- relevel(NISTEENPUF23_PDAT$BIRTH_FACTOR, ref = "2008")

# 3. Update your survey design object to include the new variable
PUF23_U13HPV_2008 <- svydesign(id = ~SEQNUMT, 
                            strata = ~STRATUM, 
                            weights = ~PROVWT_C, 
                            data = NISTEENPUF23_PDAT, 
                            nest = TRUE)

# 4. Run the model using the new factor variable
PUF23_U13HPV_2008_model <- svyglm(P_U13HPV_15INT ~ BIRTH_FACTOR, 
                   design = PUF23_U13HPV_2008, 
                   family = quasibinomial())

summary(PUF23_U13HPV_2008_model)
```
This found that 2009 was significantly higher than 2008, and 2007 was elevated (p = 0.066).
Likely that this discrepancy from previous reports is because they used a different stats approach...

2. Tracking the 2008 cohort since their 13yo data (2021)
```{r}
#DATASET: NISTEEN_08_PDAT_MLHVARS_ALL
#VARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).


MLHvars12726 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "STRATUM")

NISTEENPUF23_08_PDAT <- NISTEENPUF23[NISTEENPUF23$AGE == 15 & NISTEENPUF23$PDAT2 == 1, MLHvars12726 ] 
NISTEENPUF22_08_PDAT <- NISTEENPUF22[NISTEENPUF22$AGE == 14 & NISTEENPUF22$PDAT2 == 1, MLHvars12726 ] 
NISTEENPUF21_08_PDAT <- NISTEENPUF21[NISTEENPUF21$AGE == 13 & NISTEENPUF21$PDAT2 == 1, MLHvars12726 ] 

#combine into one dataset
NISTEEN_08_PDAT_MLHVARS_ALL <- rbind(NISTEENPUF23_08_PDAT,NISTEENPUF22_08_PDAT, NISTEENPUF21_08_PDAT)

# Divide weights by 3 to obtain the revised weights for the 3-year average [6]
NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT <- NISTEEN_08_PDAT_MLHVARS_ALL$PROVWT_C / 3

#DATASET: NISTEEN_08_PDAT_MLHVARS_ALL
#VARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).

# We multiply the 0/1 status by the weight to get the weighted success count
P_UTDHPV_15INT_2008_weighted_sum <- tapply(NISTEEN_08_PDAT_MLHVARS_ALL$P_UTDHPV_15INT * NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT, 
                           NISTEEN_08_PDAT_MLHVARS_ALL$YEAR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEEN_08_PDAT_MLHVARS_ALL_total_weighted_pop <- tapply(NISTEEN_08_PDAT_MLHVARS_ALL$NEWWT, 
                             NISTEEN_08_PDAT_MLHVARS_ALL$YEAR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_UTDHPV_15INT_08 <- (P_UTDHPV_15INT_2008_weighted_sum / NISTEEN_08_PDAT_MLHVARS_ALL_total_weighted_pop) * 100


#4 create bar graph
barplot(P_UTDHPV_15INT_08, 
        main = "2021-2023: % OF 2008 (Inferred) UTD", 
        xlab = "Survey Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 50%
```
Note that this has much higher rates that from above chart...First chart used a retrospective flag in the 2023 data, 
whereas this pulled from each survey year. 



3. % of 16yo UTD by birth year: this is where we would add the 2008 cohort.
```{r}
#DATASET: NISTEEN_16_PDAT_MLHVARS_ALL
#vARIABLE OF INTEREST: P_UTDHPV_15INT = 3+ HPV , or 2+ HPV with first shot given before age 15 and interval between first and second shot at least 5 months (minus 4 days).


MLHvars1826 <- c("SEQNUMT", "AGE", "PDAT2", "SEX", "YEAR", "PROVWT_C", "P_N13HPV", "P_NUMHPV", "P_U13HPV", "P_UTDHPV", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3C_15INT")

NISTEENPUF23_16_PDAT <- NISTEENPUF23[NISTEENPUF23$AGE == 16 & NISTEENPUF23$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF22_16_PDAT <- NISTEENPUF22[NISTEENPUF22$AGE == 16 & NISTEENPUF22$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF21_16_PDAT <- NISTEENPUF21[NISTEENPUF21$AGE == 16 & NISTEENPUF21$PDAT2 == 1, MLHvars1826 ] 
NISTEENPUF20_16_PDAT <- NISTEENPUF20[NISTEENPUF20$AGE == 16 & NISTEENPUF20$PDAT2 == 1, MLHvars1826 ]
NISTEENPUF19_16_PDAT <- NISTEENPUF19[NISTEENPUF19$AGE == 16 & NISTEENPUF19$PDAT2 == 1, MLHvars1826 ] 

#combine into one dataset
NISTEEN_16_PDAT_MLHVARS_ALL <- rbind(NISTEENPUF23_16_PDAT,NISTEENPUF22_16_PDAT, NISTEENPUF21_16_PDAT, NISTEENPUF20_16_PDAT, NISTEENPUF19_16_PDAT)

# Divide weights by 5 to obtain the revised weights for the 5-year average [6]
NISTEEN_16_PDAT_MLHVARS_ALL$NEWWT <- NISTEEN_16_PDAT_MLHVARS_ALL$PROVWT_C / 5

# Create inferred birthyear variable
NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR <- NISTEEN_16_PDAT_MLHVARS_ALL$YEAR-16

# We multiply the 0/1 status by the weight to get the weighted success count
P_UTDHPV_15INT_weighted_sum <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$P_UTDHPV_15INT * NISTEEN_16_PDAT_MLHVARS_ALL$NEWWT, 
                           NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                           sum, na.rm = TRUE)

# B. Calculate the total weighted population per birth year (Denominator)
NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop <- tapply(NISTEEN_16_PDAT_MLHVARS_ALL$NEWWT, 
                             NISTEEN_16_PDAT_MLHVARS_ALL$INFBIRTHYR, 
                             sum, na.rm = TRUE)

# C. Final Weighted Percentage Calculation (Ratio of weighted sums)
P_UTDHPV_15INT_by_16 <- (P_UTDHPV_15INT_weighted_sum / NISTEEN_16_PDAT_MLHVARS_ALLT_total_weighted_pop) * 100


#4 create bar graph
barplot(P_UTDHPV_15INT_by_16, 
        main = "2019-2023: Percentage of Teens UTD by 16yo", 
        xlab = "Inferred Birth Year", 
        ylab = "Percentage Up-to-Date (%)", 
        col = "steelblue",
        border = "white",
        ylim = c(0, 100)) # Sets the axis from 0% to 100%
```





